---
output: github_document
---

<style>
body {
text-align: justify}
</style>

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# bayesianTiterCalc

<!-- badges: start -->
[![R-CMD-check](https://github.com/ekamau/bayesianTiterCalc/workflows/R-CMD-check/badge.svg)](https://github.com/ekamau/bayesianTiterCalc/actions)
[![codecov](https://codecov.io/gh/ekamau/bayesianTiterCalc/branch/main/graph/badge.svg?token=PgQrYAAQBz)](https://codecov.io/gh/ekamau/bayesianTiterCalc)

<!-- badges: end -->

bayesianTiterCalc is a Bayesian inference method that calculates a serum sample's antibody concentration and titer based on data from a standard neutralization assay. The method uses a logistic function in a statistical model to simulate mortality in cell culture as a function of antibody concentration ($\phi$). Given the limited information per sample that's inherent in the experimental design of neutralization assays (number of replicates per dilution and number of dilutions), the advantage of Bayesian inference here is the use of probability distributions to incorporate uncertainty in the outcome.


## Installation

You can install the development version of bayesianTiterCalc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ekamau/bayesianTiterCalc")
```

## Usage

```{r setup}
library(bayesianTiterCalc)
```

The following example shows the use of the package with serum neutralization data that is provided in the package. The data was obtained from 2-fold serum dilutions from 1:8 to 1:1024, where each dilution was tested in duplicate.

```{r, message=FALSE, warning=FALSE, results='hide'}
library(dplyr)

delayedAssign('df', evTitrations)
df <- df %>% mutate(sample=as.numeric(as.factor(sampleID)))
table(df$outcome)
non_monotone <- df %>%
  group_by(sample) %>%
  summarise(is_non_monotone=non_monotonic(outcome))

table(non_monotone$is_non_monotone)

```

Visualize the data to show the number or proportion of replicates that have surviving cells (Y axis) per dilution:

```{r, out.width='70%', out.height='60%', warning=FALSE, dpi=300}
library(ggplot2)
options(dplyr.summarise.inform = FALSE)

df %>%
  filter(sample %in% 1:4) %>%
  group_by(sample, dilutions) %>%
  summarise(outcome=mean(outcome) / 2) %>%
  ggplot(aes(x=dilutions, y=outcome)) +
    geom_point() +
    scale_x_log10() +
    theme(axis.title = element_text(size=10),
          axis.text = element_text(size=8),
          strip.text = element_text(size=8)) +
    facet_wrap(~sample)

```

Fit the mechanistic model (a modified logistic function) to the data to estimate antibody concentration ($\phi$) and the endpoint serum dilution at 50\% cell culture mortality. <br>
Here we use the Stan's *sampling* algorithm, but the *optimization* algorithm is also available through the *'optimizing_stan'* function.

```{r, warning=FALSE, results='hide'}
stan_data <- list(
  N = nrow(df),
  nreplicates=rep(2, nrow(df)),
  survival=df$outcome,
  dilution=df$dilutions,
  nsample=max(df$sample),
  sample=df$sample,
  is_log=1
)

fit <- sampling_stan(standata = stan_data, chains=4, iter = 1000, init = 'random')

```

Examine the model estimates - $\phi$ and ED50 - and compare with antibody titer calculated with the Reed and Muench method. 

```{r, figures-side, fig.show='hold', out.width='50%', out.height='70%', warning=FALSE, dpi=300}
phiEstimated <- apply(rstan::extract(fit, "phi")[[1]], 2, mean)
ed50Estimated <- apply(rstan::extract(fit, "ed50")[[1]], 2, mean)
phis_ed50_titers <- df

for(i in 1:nrow(phis_ed50_titers)){
  phis_ed50_titers[i, 'phi'] <- phiEstimated[phis_ed50_titers$sample[i]];
  phis_ed50_titers[i, 'ed50'] <- ed50Estimated[phis_ed50_titers$sample[i]]
}

plot(log10(phis_ed50_titers$titer), phis_ed50_titers$phi, 
     main="Titer vs. phi", ylab="phi", xlab="log10 Titer", pch=19)
plot(log10(phis_ed50_titers$titer), phis_ed50_titers$ed50, 
     main="Titer vs. ED50", ylab="ED50", xlab="log10 Titer", pch=19)
plot(phis_ed50_titers$phi, phis_ed50_titers$ed50, 
     main="phi vs. ED50", ylab="ED50", xlab="phi", pch=19)

```

<br>
Extract posterior predictive simulations:

``` {r, results=FALSE}
survival_sim <- rstan::extract(fit, "survival_sim")[[1]]
outcome_sim <- apply(survival_sim, 2, function(x) quantile(x, 0.5))
prob <- apply(rstan::extract(fit, "prob")[[1]], 2, median)
prob_low <- apply(rstan::extract(fit, "prob")[[1]], 2, function(x) quantile(x, 0.025))
prob_high <- apply(rstan::extract(fit, "prob")[[1]], 2, function(x) quantile(x, 0.975))
df <- df %>% 
  mutate(outcome_sim = outcome_sim, prob = prob, prob_low = prob_low, prob_high = prob_high)
df_long <- df %>% 
  select(sampleID, dilutions, outcome, outcome_sim, prob, prob_low, prob_high) %>% 
  rename(actual = outcome, simulated = outcome_sim) %>% 
  mutate(fraction = actual / 2)

```

``` {r, results=FALSE, message=FALSE}
bars <- df_long %>% 
  group_by(sampleID) %>% 
  summarise(barcode = paste0(actual, collapse = "")) %>% 
  ungroup() %>% 
  pull(barcode) %>% 
  unique()
bars_lookup <- df_long %>% 
  group_by(sampleID) %>% 
  summarise(barcode = paste0(actual, collapse = ""))
df_long <- df_long %>% left_join(bars_lookup)
df_each_code <- df_long %>% 
  group_by(barcode) %>% 
  summarise(sampleID = first(sampleID)) %>% 
  mutate(keep = TRUE)
df_short <- df_long %>% 
  left_join(df_each_code) %>% 
  filter(keep == TRUE)

```

<br>
Plot the probability of cells surviving per serum dilution point: 
``` {r, message=FALSE, out.width='80%', out.height='70%', warning=FALSE, dpi=300}
library(tidyverse)
tmp <- df_short %>% 
  group_by(sampleID) %>% 
  summarise(first_non_total = which(actual < 2)[1]) %>% 
  mutate(first_non_total = ifelse(is.na(first_non_total), 
                                max(first_non_total, na.rm = T) + 1, 
                                first_non_total))
df_short %>% 
  left_join(tmp) %>% 
  mutate(sampleID = as.factor(sampleID)) %>% 
  mutate(sampleID = fct_reorder(sampleID, first_non_total)) %>%
  ggplot(aes(x = dilutions, y = prob)) +
    geom_ribbon(aes(ymin = prob_low, ymax = prob_high), alpha = 0.7, fill = "blue") +
    geom_point(aes(y = fraction), size = 0.8, alpha = 0.7, colour = "orange") +
    geom_line(aes(y = fraction), alpha = 0.7, colour = "orange") +
    scale_color_brewer("Type", palette = "Dark2") +
    scale_x_log10() +
    labs(y = "Probability of survival", x = "Dilution") +
    theme(axis.title = element_text(size=10),
          axis.text = element_text(size=8),
          strip.text = element_text(size=8)) +
    facet_wrap(~sampleID)
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
