---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bayesianTiterCalc

<!-- badges: start -->
<!-- badges: end -->

bayesianTiterCalc is a Bayesian inference method that calculates a serum sample's antibody concentration and titer based on data from a standard neutralization assay. The method uses a logistic function in a statistical model to simulate mortality in cell culture as a function of antibody concentration. Given the limited information per sample that's inherent in the experimental design of neutralization assays (number of replicates per dilution and number of dilutions), the advantage of Bayesian inference here is the use of probability distributions to incorporate uncertainty in the outcome. 


## Installation

You can install the development version of bayesianTiterCalc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ekamau/bayesianTiterCalc")
```

## Usage

What to do with the package
The most common output of a serum neutralization assay is the calculation of endpoint antibody titer. For this, 
(1) How to calculate endpoint antibody titer and concentration
(2) How to simulate data
(3) what other output do we need?

Example of data simulation: 30 serum samples with two replicates per dilution point.

```{r example}
library(bayesianTiterCalc)
phi_Vals <- c(); ndraws = 30; a = 8.5; b = 2.5; 
prior_phi <- list(n=0.75, m=16)
dilutions <- 2^c(3, 4, 5, 6, 7, 8, 9, 10)

simData <- sample_dose_response(ndraws, prior_phi, a, b, dilutions, 
                                nreplicates_per_dilution=2, phi_Vals)
head(simData)
table(simData$number_surviving)

```

The function *sample_dose_response* samples from a uniform distribution \{a, b\} and calls the *simulate_dose_response* function.

Plot four simulated samples to see how they look like and if they resemble the actual data:

```{r }
simData.plot(simData, 4)

```

Check if data is non monotonic, i.e., mortality of cell cultures increases with higher serum dilutions. Non-monotonic data would show 'accidental' death or survival of cell cultures in between dilution levels. In the *table* output, '0' and '1' represents monotonic and non-monotonic data points, respectively.

```{r }
non_monotone <- check_monotone(simData)
table(non_monotone$is_non_monotone)
```

Prepare data for analysis with Stan:

``` {r }
stan_data <- list(
  N = nrow(simData),
  nreplicates = rep(2, nrow(simData)),
  survival = simData$number_surviving,
  dilution = simData$dilution,
  nsample = max(simData$draw),
  sample = simData$draw,
  is_log = 1
)

```

Call Stan:

``` {r }
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE, javascript=FALSE)

```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
